---
title: "Gov 50 Final Project"
author: "Ella Trembanis"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project Proposal

Does school type impact to student dropout rates? Using publicly-available data on recent high school dropout rates in Delaware, I intend to assess the relationship between school type (either charter or public) and dropping out of high school. My hypothesis is that students who have both self-selected to attend charter schools, perhaps indicating a concern about the quality of public schools or a desire for a more specialized experience (such as an art- or STEM-focused charter school) and have been selected to attend, which for some charter schools involves nominally merit-based factors, such as student interviews, will be less likely to drop out. In this project, the explanatory variable is school type, designated either "Charter" or "Public". Since this variable was not present in the original dataset, a designation will be applied to each unique school code, and for the purposes of analysis, alternatives such as vocational-technical and special education schools will be excluded. The outcome variable is the student dropout rate in a given year (calculated by dividing the number of dropouts by the total student population), and will be grouped by school type. A pattern of consistent and substantially higher dropout rates among public school students would support my initial hypothesis. In mathematical terms, this would be a significant, positive difference-in-means, when calculated as the average public school dropout rate minus the average charter school dropout rate.

## Data Visualization

```{r setup}
library(tidyverse)
library(ggplot2)
dropout <- read.csv("data/Student_Dropout.csv")
```

```{r school type scatter}
dropout_type_year <- dropout |>
  mutate(
    school_type = case_when(School.Code == 26 | School.Code == 66 | School.Code == 80 | School.Code == 652 | School.Code == 770 ~ "Vocational-Technical",
                            School.Code == 354 | School.Code == 522 | School.Code == 537 | School.Code == 655 | School.Code == 689 ~ "Public (Alternative/SpecEd)",
                            School.Code == 567 | School.Code == 570 | School.Code == 574 | School.Code == 576 | School.Code == 578 | School.Code == 587 | School.Code == 4055 | School.Code == 4060 | School.Code == 4075 | School.Code == 4080 ~ "Charter",
                            School.Code == 4040 ~ "Dual Enrollment",
                            School.Code == 0 ~ "State of Delaware",
                            TRUE ~ "Public")
  ) |>
  filter(
    school_type == "Public" | school_type == "Charter"
  ) |>
  group_by(school_type, School.Year) |>
  summarize(
    avg_dropout_pct = mean(PctDropouts, na.rm = TRUE)
  )

dropout_type_scatter <- ggplot(data = dropout_type_year, mapping = aes(x = School.Year, y = avg_dropout_pct)) +
  geom_point(aes(color = school_type)) +
  geom_smooth(aes(color = school_type), span = 1) +
  labs(
    x = "School Year",
    y = "Average Dropout Rate (%)",
    title = "Yearly Dropout Rates by School Type, 2015-2022"
  )
dropout_type_scatter
```

```{r school type bar}
dropout_type <- dropout |>
  mutate(
    school_type = case_when(School.Code == 26 | School.Code == 66 | School.Code == 80 | School.Code == 652 | School.Code == 770 ~ "Vocational-Technical",
                            School.Code == 354 | School.Code == 522 | School.Code == 537 | School.Code == 655 | School.Code == 689 ~ "Public (Alternative/SpecEd)",
                            School.Code == 567 | School.Code == 570 | School.Code == 574 | School.Code == 576 | School.Code == 578 | School.Code == 587 | School.Code == 4055 | School.Code == 4060 | School.Code == 4075 | School.Code == 4080 ~ "Charter",
                            School.Code == 4040 ~ "Dual Enrollment",
                            School.Code == 0 ~ "State of Delaware",
                            TRUE ~ "Public")
  ) |>
  filter(
    school_type == "Public" | school_type == "Charter"
  ) |>
  group_by(school_type) |>
  summarize(
    avg_dropout_pct = mean(PctDropouts, na.rm = TRUE)
  )

dropout_type_bar <- ggplot(data = dropout_type, mapping = aes(x = school_type, y = avg_dropout_pct)) +
  geom_col(fill = "dodgerblue1") +
  labs(
    x = "School Type",
    y = "Average Dropout Rate (%)",
    title = "Dropout Rates by School Type, 2015-2022"
  )
dropout_type_bar
```

```{r school type diff in means}
dropout_type |>
  pivot_wider(
    names_from = school_type,
    values_from = avg_dropout_pct
  ) |>
  mutate(
    dim = `Charter` - `Public`
  ) |>
  rename("Difference in Means" = dim)
```

## Additional Plots

```{r gender scatter}
dropout_gender <- dropout |>
  filter(Gender != "All Students") |>
  group_by(Gender, School.Year) |>
  summarize(
    avg_dropout_pct = mean(PctDropouts, na.rm = TRUE)
  )

dropout_gender_scatter <- ggplot(dropout_gender, 
                             mapping = aes(x = School.Year, y = avg_dropout_pct)) +
  geom_point(aes(color = Gender)) +
  geom_smooth(aes(color = Gender), se = FALSE, span = 1) +
  labs(
    x = "School Year",
    y = "Average Dropout Rate (%)",
    title = "Yearly Dropout Rates by Gender, 2015-2022"
  )
dropout_gender_scatter
```

```{r race barplot}
dropout_race <- dropout |>
  filter(Organization == "State of Delaware") |>
  filter(Race == "African American" | Race == "White" | Race == "Asian American" | Race == "Hispanic/Latino" | Race == "Multi-Racial" | Race == "Native American") |>
  group_by(Race) |>
  summarize(
    avg_pct_dropout = mean(PctDropouts, na.rm = TRUE)
  )

dropout_bar <- ggplot(dropout_race, aes(x = Race, y = avg_pct_dropout)) +
         geom_col(fill = "dodgerblue1") +
  geom_hline(yintercept = 5.0294349, linewidth = 1.5) +
  labs(
    y = "Average Dropout Rate (%)",
    title = "Dropout Rates by Race, 2015-2022"
  ) +
  annotate("text", x = 2, y = 5.35, label = "Average Overall Dropout Rate")
dropout_bar
```

```{r race scatter}
dropout_race_year <- dropout |>
  filter(Organization == "State of Delaware") |>
  filter(Race == "African American" | Race == "White" | Race == "Asian American" | Race == "Hispanic/Latino" | Race == "Multi-Racial" | Race == "Native American") |>
  group_by(School.Year, Race) |>
  summarize(
    avg_dropout_pct = mean(PctDropouts, na.rm = TRUE)
  ) |>
  drop_na(avg_dropout_pct)

dropout_race_scatter <- ggplot(dropout_race_year, mapping = 
                                 aes(x = School.Year, y = avg_dropout_pct)) +
  geom_point(aes(color = Race)) +
  geom_smooth(aes(color = Race), se = FALSE, span = 1) +
  labs(
    x = "School Year",
    y = "Average Dropout Rate (%)",
    title = "Yearly Dropout Rates by Race, 2015-2022"
  )

dropout_race_scatter
```
